A Physically Motivated Framework to Compare Merger Timescales\\ of Isolated Low- and High-Mass Galaxy Pairs Across Cosmic Time

### 2024-03-28 paper/main.tex
added text to discussion and conclusions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
Outline draft - 
1 - Galaxy mergers are an importance feature of LCDM theory and are critical for the formation and evolution of galaxies across time. 
    - An important feature of galaxy mergers is the merger timescale...

2 - Merger timescales are important to calibrate 
    - They are crucial for understanding the observations of galaxy morphologies and understanding heirarchical evolution, (think galaxy start bursts, formation of gas bridges maybe, fueling agn, BH merger rates, etc)
    - Also important for understanding the few well modelled orbits that we do have, particularly in the LG (might come way later)

3 - Merger timescales are also used to get merger rates, both observationally and theoretically, of close pairs.
    - How are close pairs defined? 
    - How do merger timescales enter merger rate calculations?
    - What is important about the merger rates that are used? 
    - Maybe need to make distinction here that there are also observability timescales, but they are going to be basically useless for TNG100 studies (see below), but also are low mass pairs have a lot of gas and look super weird. 
    - RG 2019 looked at the simulated morphology of illustristng galaxies and found that for $M*<1e9.5$, the morphologies are "less reliable" $\to$ motivates using pair fractions for merger rate studies

4 - Comparing the merger rates of high mass and low mass pairs is important why?
    - Is important for our understanding of the build up of higher mass things in LCDM, and thus a test of LCDM
    - Different merger mechanics would be really interesting, i.e. if the DM halo profile (cuspy/core etc) impacted the merger processes

5 - Merger timescales have been studied a lot (mostly in obs and theory of high mass pairs)
    - Lotz 2011
    - RG2015
    - Snyder (maybe?) 
    - Jiang 2008 \& BK 2008 for all things about the "Chandra" timescale for merging galaxies which scales with rc/Vc and is thus prop to the age of the universe (eq. 4\&5). Also quantified the circularity of orbits as a function of mass ratio $\leftarrow$ might be good to mention
    - Jiang 2014

6 - Problem is, we don't have a lot of studies on low mass pair merger rates? 
    - This is not ideal because it means that we typically extend merger rates calibrated for high mass down to low mass, but this means that we aren't able to test LCDM and draw conclusions about the impact

@@

At each snapshot, low-mass and high-mass pairs are chosen by first selecting the two most massive subhalos (by assigned stellar mass, using abundance matching; see below) from FoF groups with virial mass 
% GB added M_G
($\rm M_{G}$)\footnote{We use \texttt{Group\_M\_TopHat200} from the TNG100 Group Catalogs as the FoF Group virial mass. This mass is defined to be the mass enclosed by a sphere with mean density $\Delta_c *\rho_c$, where $\Delta_c$ is the overdensity constant from~\citet{Bryan1998} and $\rho_c$ is the critical density of the universe at the time calculated. The corresponding virial radius in TNG100 is given by \texttt{Group\_R\_TopHat200}.} 
\begin{align*}
        \mbox{\textbf{low mass:}}&\,\rm M_{G} = 8\times 10^{10}- 5\times 10^{11}\,\Msun \\ 
        \mbox{\textbf{high mass:}}&\, \rm M_{G}=1\times 10^{12}- 6.5\times10^{12}\,\Msun.
\end{align*}
Utilizing the most massive subhalos from the same FoF group ensures that the pairs are isolated from other massive nearby systems that could perturb the dynamical state of the pair.

@@

In this analysis, a pair is classified as a merger if the primary and secondary subhalo share the same \texttt{DescendantID} at the same redshift. 
This means that the primary and secondary subhalo have merged such that \subfind{} can no longer distinguish them as two separate halos, therefore yielding a single descendant subhalo.
If a primary and secondary never share the same \texttt{DescendantID} at the same redshift, the pair is defined as a `non-merger.'

@@

The collection of all unique orbits constitutes our orbit sample, and is the dataset that will be used for the remainder of the paper.

\section{Pair Sample Properties}
\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/6_paircount.png}
    \caption{The number of low-mass (green) and high-mass (pink) unique pairs (or distinct orbits) that are post-infall and pre-merger as a function of redshift. 
    The solid lines show the total number of pairs at a given redshift (including non-mergers and mergers that have yet to merge), while dashed lines show the number of post-infall orbits at a given redshift that will merge before $z=0$.
    We only count the number of orbits ``at" a given redshift if they have not merged, but have experienced first infall at a previous time (i.e. post-infall). 
    The decrease in the number of unique low-mass pairs from $z=2$ to $z=0$ is due to pairs being removed from the sample via mergers.
    }
    \label{fig:numorbits}
    \end{center}
\end{figure}

@@

\subsection{Number of Pairs}
We calculate the number of pairs by selecting all pairs at a given redshift that have not merged.
% The number of orbits that result in mergers at a later redshift are selected in the same way, excluding the non-merging orbits.
%For example, the total number of orbits at $z=2$ includes the orbits of pairs that passed the selection criteria at \textit{any} redshift, provided they achieved first infall at $z\geq2$ and have a merger redshift $z<2$. 
A single, non-merging pair with first infall at $z=3$ will thus contribute to the number of pairs at all redshifts from $z=0-3$.

@@

\subsection{Merger Fraction}
We calculate the merger fraction by dividing the number of pairs that merge (dashed lines in Fig.~\ref{fig:numorbits}) by the number of total unique orbits (solid lines) at a given redshift, including all pairs that do and do not merge. 
As before, an orbit with first infall at $z=2$ and merger at $z=1$ will be included in the merger fraction calculation for redshifts $z=1-2$.

@@

% difference/similiarity between low and high
The merger fraction of low-mass and high-mass pairs is the same from $z=0-0.5$ and $z>2.5$. 
Between $z=0.5-2.5$, the high-mass merger fraction is higher than the low-mass merger fraction, and the knee of the merger fraction occurs at a lower redshift.
This is likely because the number of low-mass pairs continues to increase before turning over and beginning to decrease at $z\sim2$, while the number of high-mass pairs continues increasing to $z\sim1$. 
At these redshifts, the number of merging pairs begins declining rapidly. 
% When the number of pairs stops increasing and beings to decrease, pairs that will not have time to merge before $z=0$ are being added, while many merging pairs are being removed from the sample, thus decreasing the merger fraction. 
Since the turnover for the number of high-mass pairs occurs at a lower redshift, the knee of the merger fraction is likewise shifted to lower redshifts.

@@

    We find that the time until merger for low-mass and high-mass pairs is positively correlated with the separation of the pair at each redshift. 
    However, the slope of this correlation decreases at higher redshifts.
    Pairs at $z=4$ with separations between $50-100\,\kpc$ tend to merge in less time than pairs with the same separations at $z=0.5$.

@@

    Data from the table can be interpreted in the following way: at $z=1.5$, there are 5,787 pairs that are post-infall and merge prior to $z=0$. 
    Of those, 2,458 (42.47\%) have low separations between $\rsep=10-50\,\kpc$, and 2,207 of those $10-50\,\kpc$ pairs will merge in the next $2\,\Gyr$.

    \begin{figure}[htb]
        \centering
        \includegraphics[width=\columnwidth]{plots/bet-on-it/8_timescale_mod.png}
        \caption{The median time until merger as a function of redshift for low-mass (green) and high-mass (pink) pairs. Shaded regions represent the 1st and 3rd quartile spread on the median. Pairs at each redshift are selected if they have experienced first infall and are pre-merger. 
        % 
        The black dashed line shows a fraction of the Hubble time as a function of redshift, which bounds the median time until merger from $z=6$ to $z\sim0.5-0.75$.
        The dotted black line shows the time to $z=0$ (the lookback time) as a function of redshift, which sets the upper bound for the time until merger that a merging pair can have. 
        The median time until merger is similar for low-mass and high-mass pairs, and rises from $z=6$ to a peak at $z\sim0.75$, then decreases to $z=0$.\kc{only one of the dashed lines will remain in this plot once I discuss w Paul}}
        \label{fig:timescales}
    \end{figure}

@@

    First, we consider the pairs sample as a whole, and quantify the median merger timescale for all low- and high-mass pairs that are post-infall and pre-merger for $z=0-6$. 
    We will then consider two different sets of separation criteria to create separation-selected subsamples.
    Since pair samples are typically picked via separation criteria, in both simulations and observations, we study the impact of different separation-based selection criteria on inferred merger timescales.

    \subsubsection{Full sample}
        We calculate the merger timescales for pairs at each redshift in the simulation from $z=0-6$, and quantify the median and spread on the merger timescale as a function of redshift. 
        We include the full catalog of merging orbits at each redshift to calculate the merger timescale. 
        % why do we want to consider the entire population of halos? why interesting?

@@

        % redshift evolution
        The median time until merger is $\sim0.5-0.7\,\Gyr$ at $z\sim6$, then rises to a peak of $2.3-2.4\,\Gyr$ at $z\sim0.6$, then decreases to 0 at $z=0$.
        The abrupt decrease to 0 from $z\sim0.6$ to $z=0$ is artificial rather than a true physical feature. 
        As $z\to0$, there is an increasingly small fraction of pairs that merge before $z=0$, as seen by the dashed lines in Fig~\ref{fig:numorbits}. 
        These mergers must then proceed on timescales less than the elapsed time difference between a given redshift and $z=0$. 
        This is shown by the dotted black line on the left of Fig.~\ref{fig:timescales}, which shows the lookback time of the simulation as a function of redshift.\footnote{The lookback time at a given redshift $z_n$ is equivalent to the time elapsed between $z_n$ and $z=0$. Thus, merging pairs cannot have merger timescales larger than the lookback time of the simulation at a given redshift.} % this could go in the methods section, if we think it's more necessary there. I don't mind it here for more detail

    % this is all really discussion, so I can either move this to discussion and reference that we talk about the H(z) line more explicitly in the discussion, or not
        In principle, merger timescales\footnote{Specifically, the elapsed time between crossing into the virial radius of a subhalo and coalescence with the central galaxy.} for a body moving through a homogeneous field of collision-less matter can be analytically derived from the Chandrasekhar formula for dynamical friction~\citep{Binney2008}. 
        Departures from the idealized case introduce perturbations to that solution, the validity of which has been tested in cosmological hydrodynamic and N-body simulations~\citep{Jiang2008, Boylan_kolchin2008}.

@@

        In our study, we keep the stellar and FoF group mass criteria fixed as a function of redshift, such that for a given pair, the merger timescale scales with redshift as
        \begin{equation}
            \rm \tau_{merge} \propto \tau_{dyn} \propto H(z)^{-1},
        \end{equation}
        assuming that there is no (or weak) redshift dependence of the distribution of orbital parameters.
        Thus, the merger timescale is expected to scale with the Hubble Time at a given redshift.

@@

        The first set of separation criteria will select only pairs at a given redshift that have separations greater than $10\,\kpc$ and less than $[50, 70, 100, 150, 200, \mbox{and }300]\kpc$, dividing the sample into six pair subsamples. 
        These separation criteria do not vary as a function of time or mass, and are applied equivalently to the low-mass and high-mass samples at all redshifts. 
        We note that an orbit in the separation bin $10-50\,\kpc$ at $z=2$ will not necessarily be in that same bin at other redshifts. 
        Rather, the separation criteria are applied independently at each redshift.

        % figure description 
        The top panel of Fig.~\ref{fig:timescales-sep} shows the time until merger versus redshift for each of the low-mass (green) and high-mass (pink) pair subsamples with separations less than the physical separation listed above each column.  
        % redshift evolution of each subsample
        We first note that the merger timescale peaks around redshift $z\sim0.5$ for all low-mass and high-mass subsamples. 
        In addition, all subsamples show the same decline of mean merger time as $z\to0$, as discussed in the previous section. 
        These traits were also present in Fig.~\ref{fig:timescales}, so they are features that are roughly independent of a separation criteria.
        As the selection criteria gets larger, the subsample contains more of the full sample, eventually including the full sample at large enough separations.

        % low mass vs high mass
        The subsamples with maximum separations of $[50, 70, \mbox{and } 100]\,\kpc$ result in median merger timescales that are higher for low-mass pairs than high-mass pairs at all redshifts. 
        The difference in the median time until merger is up to $0.8\,\Gyr$ longer for low-mass pairs than high-mass pairs at the same redshift for the same separation criteria. 
        The offset between the low-mass and high-mass merger timescale decreases for each successive separation cut, which will each include pairs with larger separation. 
        In the rightmost top panel, the median merger times overlap almost identically for $z=0-0.5$ and $z>1.5$, almost mimicking our findings from the full sample analysis.

@@

% ##################
% Discussion section
\section{Discussion} \label{sec:discussion}
    We have calculated the merger timescales, or time until merger, for a sample of low-mass and high-mass pairs that have experienced first infall and merge before $z=0$ to ensure that they are physically collocated.
    In this section, we will explore the implications and broader impacts of our findings, and put our study in context with previous work. 
    %In Sec.~\ref{}, etc...

    \begin{itemize}
        % \item Implications for 
        \item Sec 1: Comparison to previous works: merger fraction/timescale studies
        %Why did we pick the separation cuts that we did? 
        \begin{itemize} 
            \item What other studies have looked at merger fractions/timescales in hydro simulations, specifically? 
            \item Have they studied merger fractions as a function of mass and redshift? What did they find?
            \item Were low-mass pairs expected to have such high merger rates in isolated environments? 
            \item Have they studied merger timescales as a function of mass and redshift? 
            \item What is unique about our study?
            \item Are comparisons with previous works fair, or are they not really comparable?
            \item specific papers: Ventou 2017, Hopkins 2007, O'Leary 20
        \end{itemize}

        \item Sec. 2: Implications for observational merger fractions and merger rates
        \begin{itemize}
            \item How did we motivate separation criteria?(motivated already in Chambe+ 2024, restate here about close pair definitions from lit.)
            \item Pair fractions can vary wildly depending on separation criteria
            \item This is important too for merger rates, since they are a function of pair fractions and merger timescales, and you should really be selecting pairs at each stage as self-consistently as possible
            \item What is our suggestion moving forwards?
            \item specific papers: Lotz 2011, 
        \end{itemize}
        \item Sec 3: The functional form of the merger timescale
        \begin{itemize}
            \item What can we say about the fact that the merger timescales follow a $(1+z)^{-1}$ relation at $z>0.5-1$? Physical explanation for this behaviour? 
            \item Scaled separation criteria are akin to picking the same fraction of the total volume of a FoF group, regardless of what redshift it is at, not what mass it has, so it may not be surprising that we got what we did in re: the scaled separation being better, but finding the SAME merger timescales for hi and low is interesting!
            \item Maybe something to consider in this section: For things that don't merge, what's... going on? 
            \item relevant papers: 
        \end{itemize}
    \end{itemize}

    Bring up assumption of isolation criteria and it's drawbacks and strengths.
    Reiterate that our findings are for isolated pairs.

\section{Summary and Conclusions}
    % what we did and how
    $<$Longer summary of what we did here$>$

    % set of conclusions from results
    Our main findings are as follows: 
    \begin{itemize}
        \item for a given physical separation (such as 100kpc), high-mass pairs have shorter merger timescales than low-mass pairs. 
        \item the median time until merger for the full sample of low and high-mass pairs increases as $\sim(1+z)^{-1}$ from $z=6$ to $z\sim0.75$. it then decreases to 0 at low redshifts, with an upper bound set by the time remaining in the simulation before $z=0$. The decrease in the merger time at low redshifts is artificial and not representative of the true merger timescales that pairs at $z=0$ will have.
        \item separation criteria that include larger separations will increase the median time until merger.
        \item static physical separation-selected subsamples of low- and high-mass pairs have different median merger timescales.
        \item scaled separation-selected subsamples yield nearly identical median merger timescales for low-mass and high-mass pairs at all redshifts. 
    \end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
% Outline draft - 
% 1 - Galaxy mergers are an importance feature of LCDM theory and are critical for the formation and evolution of galaxies across time. 
%     - An important feature of galaxy mergers is the merger timescale...

% 2 - Merger timescales are important to calibrate 
%     - They are crucial for understanding the observations of galaxy morphologies and understanding heirarchical evolution, (think galaxy start bursts, formation of gas bridges maybe, fueling agn, BH merger rates, etc)
    %- Also important for understanding the few well modelled orbits that we do have, particularly in the LG (might come way later)

% 3 - Merger timescales are also used to get merger rates, both observationally and theoretically, of close pairs.
%     - How are close pairs defined? 
%     - How do merger timescales enter merger rate calculations?
%     - What is important about the merger rates that are used? 
%     - Maybe need to make distinction here that there are also observability timescales, but they are going to be basically useless for TNG100 studies (see below), but also are low mass pairs have a lot of gas and look super weird.

Galaxy merger rates are a test of LCDM theory. In the era of JWST, these tests can push to higher redshifts and lower mass scales.  At high redshift and at lower masses, even isolated galaxies are highly disturbed, making pair fractions critical for merger rate studies \cite{Martin2018, RG2019}. However, there is no framework for comparing merger rates from pair fractions for low mass versus high mass galaxies and tensions exist in the literature at low masses (e.g. Vicente paper figure). i.e. There is a tension at low mass% vicente's work is done consistently, but we're basically showing that if you want to put the observational stuff on there too, you can't just willy nilly use any old pair criteria

This study seeks to establish a framework to enable comparisons of merger rates across mass scales and redshift so that merger rates of low mass galaxies can be used as tests of LCDM theory. These results can also be used to interpret observations of galaxy morphologies and their hierarchical evolution. , (think galaxy start bursts, formation of gas bridges maybe, fueling agn, BH merger rates, etc)

% 4 - Comparing the merger rates of high mass and low mass pairs is important why?
%     - Is important for our understanding of the build up of higher mass things in LCDM, and thus a test of LCDM
%     - Different merger mechanics would be really interesting, i.e. if the DM halo profile (cuspy/core etc) impacted the merger processes

% 5 - Merger timescales have been studied a lot (mostly in obs and theory of high mass pairs)
%     - Lotz 2011
%     - RG2015
%     - Snyder (maybe?) 
%     - Jiang 2008 \& BK 2008 for all things about the "Chandra" timescale for merging galaxies which scales with rc/Vc and is thus prop to the age of the universe (eq. 4\&5). Also quantified the circularity of orbits as a function of mass ratio $\leftarrow$ might be good to mention
%     - Jiang 2014

% 6 - Problem is, we don't have a lot of studies on low mass pair merger rates? 
%     - This is not ideal because it means that we typically extend merger rates calibrated for high mass down to low mass, but this means that we aren't able to test LCDM and draw conclusions about the impact

@@

At each snapshot, low-mass and high-mass pairs are chosen by first selecting the two most massive subhalos (by assigned stellar mass, using abundance matching; see below) from FoF groups with virial mass 
% GB added M_G
($\rm M_{G}$)\footnote{We use \texttt{Group\_M\_TopHat200} from the TNG100 Group Catalogs as the FoF Group virial mass. This mass is defined to be the mass enclosed by a sphere with mean density $\Delta_c *\rho_c$, where $\Delta_c$ is the overdensity constant from~\citet{Bryan1998} and $\rho_c$ is the critical density of the universe at the time calculated. The corresponding virial radius in TNG100 is given by \texttt{Group\_R\_TopHat200}.} 
\begin{align*}
        \mbox{\textbf{low mass:}}&\,\rm M_{G} = 8\times 10^{10}- 5\times 10^{11}\,\Msun \\ 
        \mbox{\textbf{high mass:}}&\, \rm M_{G}=1\times 10^{12}- 6.5\times10^{12}\,\Msun.
\end{align*}
Requiring pairs to belong to the same FoF group ensures that the pairs are isolated from other massive nearby systems that could perturb the dynamical state of the pair.

@@

In this analysis, a pair is classified as a merger if the primary and secondary subhalo share the same \texttt{DescendantID} at the same redshift. 
This means that the primary and secondary subhalo have merged such that \subfind{} can no longer distinguish them as two separate halos, therefore yielding a single descendant subhalo.
If a primary and secondary never share the same \texttt{DescendantID} at the same redshift, the pair is defined as a `non-merger.'
% in this study, we will look at the merger timescales of the merger sample.

@@

The collection of all unique orbits constitutes our orbit sample and is the dataset that will be used for the remainder of the paper.

\section{Pair Sample Properties}\label{sec:pairprops}
\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/6_paircount.png}
    \caption{The number of low-mass (green) and high-mass (pink) unique pairs (or distinct orbits) that are post-infall and pre-merger as a function of redshift. 
    The solid lines show the total number of pairs at a given redshift (including non-mergers and mergers that have yet to merge), while dashed lines show the number of post-infall orbits at a given redshift that will merge before $z=0$.
    We only count the number of orbits ``at" a given redshift if the pairs have not merged, but have experienced first infall at a previous time (i.e. post-infall). 
    The decrease in the number of unique low-mass pairs from $z=2$ to $z=0$ is due to pairs being removed from the sample via mergers.
    }
    \label{fig:numorbits}
    \end{center}
\end{figure}

@@

\subsection{Number of Pairs}\label{sec:pairprops-num}
We calculate the number of pairs by selecting all pairs at a given redshift that have not merged.
% The number of orbits that result in mergers at a later redshift are selected in the same way, excluding the non-merging orbits.
%For example, the total number of orbits at $z=2$ includes the orbits of pairs that passed the selection criteria at \textit{any} redshift, provided they achieved first infall at $z\geq2$ and have a merger redshift $z<2$. 
A single, non-merging pair with first infall at $z=3$ will thus contribute to the number of pairs at all redshifts from $z=0-3$.

@@

\subsection{Merger Fraction}\label{sec:pairprops-frac}
We calculate the merger fraction by dividing the number of pairs that merge (dashed lines in Fig.~\ref{fig:numorbits}) by the number of total unique orbits (solid lines) at a given redshift, including all pairs that do and do not merge. 
As before, an orbit with first infall at $z=2$ and merger at $z=1$ will be included in the merger fraction calculation for redshifts $z=1-2$.

@@

% difference/similiarity between low and high
The merger fraction of low-mass and high-mass pairs is the same from $z=0-0.5$ and $z>2.5$. 
Between $z=0.5-2.5$, the high-mass merger fraction is higher than the low-mass merger fraction, and the knee of the merger fraction occurs at a lower redshift.
This is likely because the number of low-mass pairs continues to increase before turning over and beginning to decrease at $z\sim2$, while the number of high-mass pairs continues increasing to $z\sim1$. 
At these redshifts, the number of merging pairs begins declining rapidly. 
% When the number of pairs stops increasing and beings to decrease, pairs that will not have time to merge before $z=0$ are being added, while many merging pairs are being removed from the sample, thus decreasing the merger fraction. 
Since the turnover for the number of high-mass pairs occurs at a lower redshift, the knee of the merger fraction is likewise shifted to lower redshifts.
% move the above to discussion and end with something shorter: 
% This COULD be that the timescales. This will answer this with this study and discuss this in detail in Disc.~\ref.

@@

    We find that the time until merger for low-mass and high-mass pairs is positively correlated with the separation of the pair at each redshift. 
    However, the slope of this correlation decreases at higher redshifts. 
    % For the same physical separation, the merger timescale is lower than for low-mass 
    % This does that the overall merger timescales are lower for high-mass systems. 
    % This is discussed in detail in the next section.
    Pairs at $z=4$ with separations between $50-100\,\kpc$ tend to merge in less time than pairs with the same separations at $z=0.5$.

@@

    Data from the table can be interpreted in the following way: at $z=1.5$, there are 5,787 pairs that merge prior to $z=0$. 
    Of those, 2,458 (42.47\%) have low separations between $\rsep=10-50\,\kpc$, and 2,207 (38.1\%) of those $10-50\,\kpc$ pairs will merge in the next $2\,\Gyr$.

    \begin{figure}[htb]
        \centering
        \includegraphics[width=\columnwidth]{plots/bet-on-it/8_timescale_mod.png}
        \caption{The median time until merger as a function of redshift for low-mass (green) and high-mass (pink) pairs that will merge by $z=0$. Shaded regions represent the 1st and 3rd quartile spread on the median. %Pairs at each redshift are selected if they have experienced first infall and are pre-merger. 
        % 
        The black dashed line shows the fraction of the Hubble time as a function of redshift, which bounds the median time until merger from $z=6$ to $z\sim0.5-0.75$.
        The dotted black line shows the time to $z=0$ (the lookback time) as a function of redshift, which sets the upper bound for the time until merger that a merging pair can have. 
        The median time until merger is similar for low-mass and high-mass pairs, and rises from $z=6$ to a peak at $z\sim0.75$, then decreases to $z=0$.\kc{only one of the dashed lines will remain in this plot once I discuss w Paul; likely remove age of sim}}
        \label{fig:timescales}
    \end{figure}

@@

    First, we consider the pairs sample as a whole, and quantify the median merger timescale for all low- and high-mass pairs that will merge for $z=0-6$. 
    We will then consider two different sets of separation criteria to create separation-selected subsamples.
    Since pair samples are typically picked via separation criteria, in both simulations and observations, we study the impact of different separation-based selection criteria on inferred merger timescales.

    \subsubsection{Full sample}
        We calculate the merger timescales for pairs at each redshift in the simulation from $z=0-6$, and quantify the median and spread on the merger timescale as a function of redshift. 
        We include the full catalog of merging orbits at each redshift.% to calculate the merger timescale. 
        % why do we want to consider the entire population of halos? why interesting?

@@

        % redshift evolution
        The median time until merger is $\sim0.5-0.7\,\Gyr$ at $z\sim6$, then rises to a peak of $2.3-2.4\,\Gyr$ at $z\sim0.6$, then decreases to 0 at $z=0$.
        The abrupt decrease is artificial rather than a true physical feature. 
        As $z\to0$, there is an increasingly small fraction of pairs that merge before $z=0$, as seen by the dashed lines in Fig~\ref{fig:numorbits}. 
        These mergers must then proceed on timescales less than the elapsed time difference between a given redshift and $z=0$. These mergers must then proceed on decreasing timescales by definition, as they were selected as pairs that must merge prior to $z=0$ (i.e., a pair chosen by the selection criteria given in Section XYZ at z=0.2 must merge by $z=0$, giving a maximal merger timescale equivalent XYZ Gyr whereas a pair chosen at z=6 can have a maximal merger timescale of XYZ Gyr.). 
        This is shown by the dotted black line on the left of Fig.~\ref{fig:timescales}, which shows the lookback time of the simulation as a function of redshift.\footnote{The lookback time at a given redshift $z_n$ is equivalent to the time elapsed between $z_n$ and $z=0$. Thus, merging pairs cannot have merger timescales larger than the lookback time of the simulation at a given redshift.} % this could go in the methods section, if we think it's more necessary there. I don't mind it here for more detail

        In principle, merger timescales\footnote{Specifically, the elapsed time between crossing into the virial radius of the primary subhalo and coalescence with the primary.} for a body moving through a homogeneous field of collision-less matter can be analytically derived from the Chandrasekhar formula for dynamical friction~\citep{Binney2008}. 
        Departures from the idealized case introduce perturbations to that solution, the validity of which has been tested in cosmological hydrodynamic and N-body simulations~\citep{Jiang2008, Boylan_kolchin2008}.

@@

        In our study, we keep the stellar and FoF group mass criteria fixed as a function of redshift, such that for a given pair, the merger timescale scales with redshift as
        \begin{equation}
            \rm \tau_{merge} \propto \tau_{dyn} \propto H(z)^{-1},
        \end{equation}
        assuming that there is no (or weak) redshift dependence of the distribution of orbital parameters.
        Thus, the merger timescale is expected to scale with the Hubble time at a given redshift.

@@

        The first set of separation criteria will select only pairs at a given redshift that have physical 3D separations greater than $10\,\kpc$ and less than $[50, 70, 100, 150, 200, \mbox{and }300]\kpc$, dividing the sample into six pair subsamples. 
        These separation criteria do not vary as a function of time or mass, and are applied equivalently to the low-mass and high-mass samples at all redshifts. 
        For example, an orbit in the separation bin $10-50\,\kpc$ at $z=2$ will not necessarily be in that same bin at other redshifts. 
        Rather, the separation criteria are applied independently at each redshift.

        % figure description 
        The top panel of Fig.~\ref{fig:timescales-sep} shows the time until merger versus redshift for each of the low-mass (green) and high-mass (pink) pair subsamples with separations less than the physical separation listed above each column.  
        % redshift evolution of each subsample
        We first note that the merger timescale peaks around redshift $z\sim0.5$ for all low-mass and high-mass subsamples. 
        In addition, all subsamples show the same decline of mean merger time as $z\to0$, as discussed in the previous section. 
        These traits were also present in Fig.~\ref{fig:timescales}, so they are features that are roughly independent of a separation criteria.
        As the selection criteria gets larger, the subsample contains more of the full sample, eventually including the full sample of merging pairs at large enough separations.

        % low mass vs high mass
        The subsamples with maximum separations of $[50, 70, \mbox{and } 100]\,\kpc$ result in median merger timescales that are higher for low-mass pairs than high-mass pairs at all redshifts. 
        The difference in the median time until merger is up to $0.8\,\Gyr$ longer for low-mass pairs than high-mass pairs at the same redshift for the same separation criteria. 
        The offset between the low-mass and high-mass merger timescale decreases for each successive separation cut, which will each include pairs with larger separation. 
        In the rightmost top panel, the median merger times converge for $z=0-0.5$ and $z>1.5$, as expected based on results from the full sample shown in Fig. \ref{fig:timescales}. %almost mimicking our findings from the full sample analysis.

@@

% ##################
% Discussion section
\section{Discussion} \label{sec:discussion}
    We have calculated the merger timescales, or time until merger, for a sample of physically collocated low-mass and high-mass pairs from $z=0-6$.
    In this section, we will explore the implications and broader impacts of our findings, and put our study in context with previous work. 
    %In Sec.~\ref{}, etc...

    In Sec.~\ref{sec:disc-nonmergers}, we will discuss the non-merger portion of our orbit sample and the difference between low- and high-mass merger fractions.
    In Sec.~\ref{sec:disc-previous}, we discuss our work in the context of other studies that have sought to reconcile xx xmerger rates and merger timescales via a change in pair selection criteria. 
    In Sec.~\ref{sec:disc-suggestions}, we discuss the implications of our work for future observational and theoretical merger fraction and merger rate studies and provide suggestions for a self-consistent framework moving forward.
    Finally, in Sec.~\ref{sec:disc-phys}, we briefly discuss the underlying physical mechanisms implied by our results and the functional form of the merger timescale.

    \subsection{Non-merger population}\label{sec:disc-nonmergers}
    In our catalog of 22,213 low-mass pairs and 3,029 high-mass pairs, only 16.66\% (3700) of low-mass pairs and 29.71\% (903) of high-mass pairs do not merge by $z=0$.
    In Sec.~\ref{sec:pairprops-frac}, we found that the merger fraction of low-mass and high-mass pairs at $z>3$ is $>0.95$. 
    This changes as you move to lower redshifts, where the merger fraction decreases substantially. 
    While the decline in the merger fraction at $z\lesssim1$ is primarily due to the limited time that a pair will have to merge before the end of the simulation, we also find that between $z=\sim0.6-2.5$ the fraction of low-mass pairs is lower than that of high-mass pairs.

    One potential explanation for this behavior could be that low-mass pairs have shorter merger timescales than high mass pairs, and thus merge at earlier times in the simulations, thus leading to a decrease in the merger fraction at higher redshifts. 
    However, through our analysis, we have shown that the merger timescales of our full low-mass and high-mass pair sample evolve in the same way from $z=0-6$. 
    This means that the difference in the merger fraction between $z=0.5-2.5$ is not a result of different merger timescales.

    We suspect that the cause of the difference between low- and high-mass merger fractions is not a difference in the merging population itself, but rather in the non-merger population and its evolution over time. 
    As we did not distinguish our non-merging sample into "likely mergers" and "fly-by" interactions for the purposes of the study, we cannot say anything about how these two populations individually contribute to the difference in the merger fraction.
    However, one possible explanation could be that a larger fraction of infalling low-mass pairs at lower redshift result in fly-by interactions. 
    This could be the case if the velocity at which secondary subhalo enter the primary's FoF group gets increasingly large compared to the virial velocity\footnote{The virial velocity is the value of the circular velocity at the virial radius.}, resulting in fewer bound systems. 
    We leave this as a focus of future work.

    % Specifically, we believe that the separation and velocity of the pairs that experience first infall at lower redshifts ($z<2.5$) lead to but do not merge in the simulation are driving the difference, and that more of the infalling low-mass pairs result in non-mergers than the high-mass pairs.
    % More specifically, if low-mass pairs at low redshift ($z<3$) have infall velocities that are lower with respect to the virial velocity than their high-mass counterparts, then the merger timescale for that sample of longer-.

    \subsection{Steps Towards Standardized Pair Selection Criteria}~\label{sec:disc-previous}
    Comparisons between studies of pair fractions, merger fractions, and merger rates across cosmic time and different mass scales are often made challenging by the use of different sets of selection criteria in each study. 
    On the observational side, pair selection criteria are often set by the observational parameters of the survey itself. 
    For example, limiting magnitudes and completeness limits can enforce a specific new range of stellar mass ratios to be considered (for example $1-1:6$ instead of $1:10$), while avoiding colliding/overlapping fibers on a spectroscopic survey might lead to very specific separation criteria.

    On the theoretical side, it can be straightforward to adopt a number of different pair selection criteria for comparisons to specific observational studies with identical criteria. 
    However, this means that theoretical studies typically restrict themselves to a subset of pair selection criteria specific to their target comparison studies.

    A variety of studies in recent years aim to standardize pair selection criteria for future studies, in particular to permit a more fair comparison between theory and observations. 
    One such study is~\citet{Ventou2017}, which created mock catalogs from the original Illustris-1 simulation to analyze the merger probability of galaxy pairs as a function of projected separation and projected relative velocity. 
    They find that 
    to provide a new set of selection criteria to define close pairs. 
    One

    Merger rates/merger fractions in illustris, close pair fractions and merger timescales, etc. 
    As in our previous work on pair fractions in TNG100, we are unable to make meaningful comparisons with previous observational studies of merger fraction and merger rates, as our study is completed using true physical separations from the simulation. 
    However, 
    % In~\cite{RG2015}, they quantify the galaxy major merger rate as a function of the stellar mass of the descendent subhalo. 
    % It is unclear, however, if the major merger rate for galaxies with stellar mass $\ms<10^{10}\Msun$ is actually discrepant between simulations and observations, or if the observational sam

    NOT a lit review of merger topics etc. 
        \begin{itemize}
            \item Have they studied merger timescales as a function of mass and redshift? 
            \item What is unique about our study?
            \item Are comparisons with previous works fair, or are they not really comparable?
            \item specific papers: Ventou 2017, Hopkins 2007, O'Leary 20
            \item only look at merger rates on the low mass end ? - garreth 
            \item studies that redefine pair fraction criteria 
            \item What other studies have looked at merger fractions/timescales in hydro simulations, specifically? 
            \item Have they studied merger fractions as a function of mass and redshift? What did they find?
            \item Were low-mass pairs expected to have such high merger rates in isolated environments? 
        \end{itemize}

    \subsection{Implications for Future Observational and Theoretical Studies of Pairs}~\label{sec:disc-suggestions}
    In Sec.~\ref{sec:results-timevredshift}, we find that the merger timescales of isolated low-mass and high-mass pairs evolve nearly identically between $z=0-6$, rising from a median of $\sim0.8Gyr$ at $z=6$ to $2.1-2.4\Gyr$ at $z\sim0.6$.
    From $z=0$ to $z=0.6$, the merger timescale rapidly declines, since the only mergers at low redshift that can proceed before the end of the simulation at $z=0$ are those with very short timescales.

    In our previous work, we studied the impact of different sets of separation criteria on the recovered pair fractions of low-mass and high-mass pairs, and found that only by utilizing scaled separation criteria can the features of the pair fractions' evolution with redshift be distinguished~\citep{Chamberlain2024}. 
    In this work, we explored how these same set of selection criteria impacts the recovered merger timescales.
